---
title: "Hands-On Tutorials for Covariate Adjustment in Randomized Trials"
subtitle: "Time-To-Event Outcome"
author: "Josh Betz (jbetz@jhu.edu), Kelly Van Lancker (kvanlan3@jhu.edu), and Michael Rosenblum (mrosen@jhu.edu)"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
header-includes:
   - \usepackage{amsmath}
output:
  ioslides_presentation:
    widescreen: true
    incremental: true
bibliography: covariate_adjustment.bib # For Bibliography - In same folder
---




```{r setup, include = FALSE}
# Packages used for formatting, not needed for analysis
library(pander)
library(knitr)
library(Cairo)

### Graphics and Report Options ################################################
table_digits <- 1 # Significant Figures for Mean/SD, N (%), Median/IQR

root_url <- "https://jbetz-jhu.github.io/CovariateAdjustmentTutorial"

load(
  url(
    paste0("https://github.com/kelvlanc/PSI2023-Covariate-Adjustment/raw/",
           "main/Betz/sim_mistie_iii_1.2_tmle.Rdata")
  )
)


bootstrap_tutorial_url <-
  paste0(root_url, "/bootstrap_tutorial.html")

estimands_tutorial_url <-
  paste0(root_url, "/blob/gh-pages/Estimands_of_Interest.html")

estimators_tutorial_url <-
  paste0(root_url, "/blob/gh-pages/Estimators.html")




### Set Default Options ########################################################
options(
  knitr.kable.NA = "",
  width = 300
)

run_adjrct <- FALSE
run_adjrct <- TRUE # Takes time


### Set Default Options ########################################################
fig_w <- 8
fig_h <- 8

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  results = "markup",
  fig.width = fig_w,
  fig.height = fig_h,
  fig.align = "center",
  out.width = "90%",
  out.height = "90%",
  dpi = 600,
  dev = "CairoPNG"
)

### options ####################################################################
options(
  knitr.kable.NA = ''
)
```

```{r Load-Data-and-Packages, echo = FALSE, message = FALSE}
suppressWarnings({
  suppressMessages({
    library(knitr) # Printing tables in reports
    library(tidyverse) # Data manipulation: dplyr, tidyr
    library(table1) # Creation of Summary tables
    library(survival) # AFT & Cox PH Model
    library(survminer) # Kaplan-Meier Plots
    library(coxrobust) # coxr: Robust Cox PH Model
    library(survRM2) # Restricted mean survival times (RMST)
    library(adjrct) # RMST, Survival Probability
    library(splines) # Smoothing splines
  })
})


data_url <-
  "https://github.com/jbetz-jhu/CovariateAdjustmentTutorial/raw/main/Simulated_MISTIE_III_v1.2.csv"

sim_miii <-
  read.csv(file = url(data_url)) %>% 
  dplyr::slice(
    1:500
  )

# Read in data: Recast categorical variables as factors
sim_miii <-
  sim_miii %>% 
  dplyr::tibble() %>% 
  dplyr::mutate(
    male =
      factor(x = male, levels = 0:1, labels = c("0. Female", "1. Male")),
    across(
      .cols = all_of(x = c("hx_cvd", "hx_hyperlipidemia",
                           "on_anticoagulants", "on_antiplatelets")),
      .fns = function(x) factor(x, levels = 0:1, labels = c("0. No", "1. Yes"))
    ),
    across(
      .cols = starts_with("gcs") | starts_with("mrs"),
      .fns = factor
    ),
    ich_location =
      factor(
        x = ich_location,
        levels = c("Deep", "Lobar")
      ),
    arm =
      factor(
        x = arm,
        levels = c("medical", "surgical")
      ),
    tx = 1*(arm == "surgical")
  )
```


<style>
.forceBreak { -webkit-column-break-after: always; break-after: column; }
</style>



## Outline:

  - Covariate Adjustment Tutorials page
  - Worked Example: Time-To-Event Outcomes
    - Example: Mortality in MISTIE III Hemorrhagic Stroke Trial
    - Unadjusted: Logrank, Cox, Survival Probability, RMST
    - Adjusted: Cox, Survival Probability, RMST




# Hands-On Tutorials for Covariate Adjustment | [https://bit.ly/covariateadjustedrcts](https://bit.ly/covariateadjustedrcts)




## Tutorials on Covariate Adjustment | [https://bit.ly/covariateadjustedrcts](https://bit.ly/covariateadjustedrcts)

  - Simulated data: mimic features of trials - Scale, correlations, missingness
    - Continuous, Ordinal, Binary, Time-to-Event outcomes: Covariates
    - Example datasets: Substance Abuse, Hemorrhagic Stroke
    - Stratified randomization: improve precision
  - Common estimands of interest; Analytic approaches
  - R code for tabulating, plotting, analyzing data:
    - Unadjusted & Adjusted
  - Links to resources on learning and using R




# Example: based on MISTIE III | Functional Outcome & Mortality in Hemorrhagic Stroke




## MISTIE-III Trial: [@Hanley2019]

  - Hemorrhagic Stroke: Greater morbidity, mortality than ischemic stroke
    - Intracerebral Hemorrhage (ICH), possibly with Intraventricular Hemorrhage (IVH)
  - Consent: monitor daily for ICH stability by CT
  - 1:1 randomized - minimally invasive surgery + thrombolytic vs. SOC medical management
  - Safety & Efficacy: Functional outcome on Modified Rankin Scale (MRS)
    - MRS at 30, 180, and 365 days post randomization
    - Mortality: Administratively censored at 1-year follow-up
  - Good Outcome: MRS 0-3 vs. 4-6 - independent vs. not
  - Simulated data based on actual trial data: **not actual study data.**




## Simulated MISTIE Data: {.columns-2 .smaller}

  - Baseline Covariates
    - `age`: Age in years
    - `male`: male sex
    - `hx_cvd`:	cardiovascular disease history
    - `hx_hyperlipidemia`:	hyperlipidemia
    - `on_anticoagulants`:	on anticoagulants
    - `on_antiplatelets`:	on antiplatelets
    - `ich_location`: ICH: (`Lobar`, `Deep`)
    - `ich_s_volume`:	ICH volume on stability scan
    - `ivh_s_volume`:	IVH volume on stability scan
    - `gcs_category`: presenting Glasgow Coma Score


<p class="forceBreak"></p>


  - Treatment:
    - `arm`: treatment arm
    - `ich_eot_volume`: intracerebral hemorrhage volume on end-of-treatment scan
  - Outcome:
    - Modified Rankin: `_complete`: completely observed
    - `mrs_30d`: MRS at 30 days (`0-3`, `4`, `5`, `6`)
    - `mrs_180d`: MRS at 180 days (`0-2`, `3`, `4`, `5`, `6`)
    - `mrs_365d`: MRS at 365 days (`0-1`, `2`, `3`, `4`, `5`, `6`)
    - `days_on_study`: days until death/censoring
    - `died_on_study`: participant died (`1`) or censored (`0`)


# Show of Hands: Familiarity with R | Beginner, Intermediate, Expert - **NO JUDGEMENT!**


## Installing R Packages from GitHub {.columns-2 .smaller}

  - "Base R" comes with set of packages:
    - Expanded to extend capabilities
  - [CRAN](https://cran.r-project.org/): Comprehensive R Archival Network
    - Can use `install_packages()` to install packages from CRAN, BioC, R-Forge
  - [RTools](https://cran.r-project.org/bin/windows/Rtools/): Tools for compiling packages from source
  - [Github](https://github.com/): developing software and documentation with version control
    - Can Use `devtools::install_github` to install packages hosted on Github
  - Load installed packages using `library()`


<p class="forceBreak"></p>


```{r install-github, eval = FALSE}
devtools::install_github("nt-williams/simul")
devtools::install_github("nt-williams/adjrct")
```

```{r load-packages, eval = FALSE}
library(knitr) # Printing tables in reports
library(tidyverse) # Data manipulation: dplyr, tidyr
library(table1) # Creation of Summary tables
library(survival) # AFT & Cox PH Model
library(survminer) # Kaplan-Meier Plots
library(coxrobust) # coxr: Robust Cox PH Model
library(survRM2) # Restricted mean survival times (RMST)
library(adjrct) # RMST, Survival Probability
library(splines) # Smoothing splines
```




# Baseline Comparability | Balance of Covariates




## Baseline Comparability  {.columns-2 .smaller}

Below are summary statistics of participant characteristics at baseline.

```{r table-baseline-example-code, eval = FALSE}
table1::table1(
  ~ age + male +
    on_antiplatelets +
    ich_location +
    ich_s_volume +
    ivh_s_volume +
    gcs_category | arm, 
  data = sim_miii
)
```


<p class="forceBreak"></p>


```{r table-baseline-evaluate, echo = FALSE}
table1::table1(
  ~ age + male +
    on_antiplatelets +
    ich_location +
    ich_s_volume +
    ivh_s_volume +
    gcs_category | arm, 
  data = sim_miii
)
```




## Conventional Unadjusted Estimators: {.smaller}

  - Kaplan Meier: assumes censoring independent of event time in treatment arm
    - Violated: baseline covariates associated with event times & dropout [@Diaz2018]
  - Logrank Test: valid if censoring is independent of treatment or the event time in each treatment arm [@VanLancker2021].
    - Most powerful under proportional hazards assumption: weighted logrank test [@Lin2017]
  - Cox Proportional Hazards (PH) Model: Score test equivalent to Logrank Test
    - Valid test of null: Sandwich estimator is used and either:
      - Censoring conditionally independent of treatment assignment given the covariates
      - Censoring conditionally independent of the covariates given treatment assignment
    - PH Assumption violated: weighted average of hazards, depend on censoring distribution
    - Hazard: time-varying instantaneous rate of events occur: 
      - Does not quantify the expected length of time until the event [@Rudser2012]




## Alternatives to Logrank & Cox PH

  - Time Horizon: specific, clinically meaningful time to assess outcome: $\tau$
  - Restricted Mean Survival Time: average time-to-event (e.g. life expectancy when mortality is the event of interest) from baseline to a pre-specified point in time, denoted $\tau$.
  - Survival Probability (SP): probability of being event-free at a pre-specified point in time, denoted $\tau$.
  - SP and RMST interpretable when PH assumption is violated: Clinically relevant
  - Doubly-robust estimators for SP and RMST and estimators exist - at least as efficient as the K-M estimator [@Diaz2018].




## Kaplan Meier {.columns-2 .smaller}

```{r kaplan-meier-death-example-code, eval = FALSE}
time_to_death_km <-
  survfit(
    formula =
      Surv(time = days_on_study,
           event = died_on_study) ~ arm,
    data = sim_miii
  )

ggsurvplot(
  fit = time_to_death_km,
  conf.int = TRUE,
  risk.table = TRUE,
  xlab = "Days", 
  ylab = "Survival probability"
)
```


<p class="forceBreak"></p>


```{r kaplan-meier-death-evaluate, echo = FALSE}
time_to_death_km <-
  survfit(
    formula =
      Surv(time = days_on_study,
           event = died_on_study) ~ arm,
    data = sim_miii
  )

ggsurvplot(
  fit = time_to_death_km,
  conf.int = TRUE,
  risk.table = TRUE,
  xlab = "Days", 
  ylab = "Survival probability"
)
```




## Logrank Test

  - Logrank test (G-rho family of tests): `survival::survdiff`

```{r unadjusted-logrank-test-evaluate}
survival::survdiff(
  formula =
    Surv(time = days_on_study,
         event = died_on_study) ~ arm,
  data = sim_miii
)
```




## Unadjusted Cox Model: {.smaller}

```{r unadjusted-Cox-PH}
unadjusted_cox <-
  survival::coxph(
    formula =
      Surv(time = days_on_study,
           event = died_on_study) ~ arm,
    ties = "efron",
    robust = TRUE,
    data = sim_miii
  )
summary(unadjusted_cox)
```




## Proportional Hazards Assumption {.columns-2 .smaller}

```{r unadjusted-Cox-PH-test-echo, eval = FALSE}
unadjusted_cox_ph_test <-
  cox.zph(unadjusted_cox)
print(unadjusted_cox_ph_test)
# Plot Proportionality Results
plot(cox.zph(unadjusted_cox),
     main = "Proportional Hazards Test")
abline(h = 0, col = "red")
```

```{r unadjusted-Cox-PH-test-evaluate, echo = FALSE}
unadjusted_cox_ph_test <-
  cox.zph(unadjusted_cox)
print(unadjusted_cox_ph_test)
```


<p class="forceBreak"></p>

```{r unadjusted-Cox-PH-plot-evaluate, echo = FALSE}
plot(cox.zph(unadjusted_cox),
     main = "Proportional Hazards Test")
abline(h = 0, col = "red")
```




## Unadjusted Restricted Mean Survival Time {.columns-2 .smaller}

  - RMST in first 90 days:
  - ~85 days in the surgical arm
  - ~76 days in the medical arm
  - Difference: 8.8 days
  - Ratio: 1.12 (12% increase in RMST).

```{r unadjusted-restricted-mean-survival-time-rmst2-echo, eval = FALSE}
with(
  sim_miii,
  survRM2::rmst2(
    time = days_on_study,
    status = died_on_study,
    arm = tx,
    tau = 90
  )
)
```


<p class="forceBreak"></p>


```{r unadjusted-restricted-mean-survival-time-rmst2-evaluate, echo = FALSE}
with(
  sim_miii,
  survRM2::rmst2(
    time = days_on_study,
    status = died_on_study,
    arm = tx,
    tau = 90
  )
)
```




## TMLE: RMST & Survival Probability {.columns-2 .smaller}

  - Targeted Maximum Likelihood Estimation
  - Compute "survival metadata" using `survrct`
  - Use result to compute RMST & Survival Probability


<p class="forceBreak"></p>


```{r unadjusted-tmle-survival-metadata-adjrct, eval = FALSE}
surv_metadata_unadj <-
  adjrct::survrct(
    outcome.formula =
      Surv(days_on_study, died_on_study) ~ tx, 
    trt.formula = tx ~ 1,
    data = sim_miii,
  )
```

```{r unadjusted-tmle-survival-rmst-adjrct, eval = FALSE}
adjrct::rmst(
  metadata = surv_metadata_unadj,
  horizon = 90
)
```

```{r unadjusted-tmle-survival-probability-adjrct, eval = FALSE}
adjrct::survprob(
  metadata = surv_metadata_unadj,
  horizon = 90
)
```




## TMLE: RMST & Survival Probability {.columns-2 .smaller}

```{r echo = FALSE}
rmst_tmle_unadj <-
  adjrct::rmst(
    metadata = surv_metadata_unadj,
    horizon = 90
  )

rmst_metadata_unadj_table <-
  with(
    rmst_tmle_unadj$estimates[[1]],

    bind_rows(
      data.frame(
        Arm = "Treatment",
        Estimate = arm1,
        SE = arm1.std.error,
        LCL = arm1.conf.low,
        UCL = arm1.conf.high
      ),

      data.frame(
        Arm = "Control",
        Estimate = arm0,
        SE = arm0.std.error,
        LCL = arm0.conf.low,
        UCL = arm0.conf.high
      ),
      data.frame(
        Arm = "Treatment - Control",
        Estimate = theta,
        SE = std.error,
        LCL = theta.conf.low,
        UCL = theta.conf.high
      )
    )
  )

kable(
  x = rmst_metadata_unadj_table,
  caption = "TMLE unadjusted estimates of Restricted Mean Survival Time.",
  digits = 2
)
```


<p class="forceBreak"></p>


```{r echo = FALSE, warning = FALSE}
survival_probability_unadj <-
  adjrct::survprob(
    metadata = surv_metadata_unadj,
    horizon = 90
  )

survival_probability_unadj_table <-
  with(
    survival_probability_unadj$estimates[[1]],

    bind_rows(
      data.frame(
        Arm = "Treatment",
        Estimate = arm1,
        SE = arm1.std.error,
        LCL = arm1.conf.low,
        UCL = arm1.conf.high
      ),

      data.frame(
        Arm = "Control",
        Estimate = arm0,
        SE = arm0.std.error,
        LCL = arm0.conf.low,
        UCL = arm0.conf.high
      ),
      data.frame(
        Arm = "Treatment - Control",
        Estimate = theta,
        SE = std.error,
        LCL = theta.conf.low,
        UCL = theta.conf.high
      )
    )
  )

kable(
  x = survival_probability_unadj_table,
  caption = "TMLE unadjusted estimates of Restricted Mean Survival Time.",
  digits = 2
)
```




# Covariate Adjusted Analyses | Cox PH, RMST, Survival Probability



## Adjusted Cox PH Model

```{r adjusted-cox-linear}
adjusted_cox_linear <-
  coxph(
    formula = Surv(days_on_study, died_on_study) ~ arm +
      # Covariates
      age + male + hx_cvd + hx_hyperlipidemia +
      on_anticoagulants + on_antiplatelets + ich_location +
      ich_s_volume + ivh_s_volume + gcs_category,
    ties = "efron", # Efron's Method for Ties
    robust = TRUE, # Use Robust SEs
    data = sim_miii
  )
```




## Adjustment with Penalized Splines: `survival::pspline()`

```{r adjusted-cox-pspline}
adjusted_cox_ps <-
  coxph(
    formula = Surv(days_on_study, died_on_study) ~ arm +
      # Covariates
      pspline(age) + male + hx_cvd + hx_hyperlipidemia +
      on_anticoagulants + on_antiplatelets + ich_location +
      pspline(ich_s_volume) + pspline(ivh_s_volume) + gcs_category,
    # Use Efron's Method for Tied Event Times
    ties = "efron",
    # Use Robust Standard Errors
    robust = TRUE,
    data = sim_miii
  )
```




## Adjusted Cox Proportionality {.columns-2 .smaller}

```{r adjusted-cox-ps-proportionality}
cox.zph(
  fit = adjusted_cox_ps
)
```


<p class="forceBreak"></p>


```{r adjusted-cox-ps-proportionality-plot}
cox.zph(fit = adjusted_cox_ps) %>% 
  plot(var = "arm")
abline(h = 0, col = rgb(1, 0, 0, 0.75))
```




## Adjusted TMLE: RMST & Survival Probability {.smaller}

```{r adjusted-tmle-metadata, eval = FALSE}
surv_metadata_adj <-
  adjrct::survrct(
    outcome.formula = Surv(days_on_study, died_on_study) ~ tx +
      age + male + hx_cvd + hx_hyperlipidemia +
      on_anticoagulants + on_antiplatelets + ich_location +
      ich_s_volume + ivh_s_volume + gcs_category,
    trt.formula =
      tx ~
      age + male + hx_cvd + hx_hyperlipidemia +
      on_anticoagulants + on_antiplatelets + ich_location +
      ich_s_volume + ivh_s_volume + gcs_category,
    data = sim_miii
  )
```


```{r adjusted-tmle-survival-rmst-adjrct, eval = FALSE}
adjrct::rmst(
  metadata = surv_metadata_adj,
  horizon = 90
)
```


```{r adjusted-tmle-survival-probability-adjrct, eval = FALSE}
adjrct::survprob(
  metadata = surv_metadata_adj,
  horizon = 90
)
```




## TMLE: RMST & Survival Probability {.columns-2 .smaller}

```{r echo = FALSE}
rmst_tmle_adj <-
  adjrct::rmst(
    metadata = surv_metadata_adj,
    horizon = 90
  )

rmst_metadata_adj_table <-
  with(
    rmst_tmle_adj$estimates[[1]],

    bind_rows(
      data.frame(
        Arm = "Treatment",
        Estimate = arm1,
        SE = arm1.std.error,
        LCL = arm1.conf.low,
        UCL = arm1.conf.high
      ),

      data.frame(
        Arm = "Control",
        Estimate = arm0,
        SE = arm0.std.error,
        LCL = arm0.conf.low,
        UCL = arm0.conf.high
      ),
      data.frame(
        Arm = "Treatment - Control",
        Estimate = theta,
        SE = std.error,
        LCL = theta.conf.low,
        UCL = theta.conf.high
      )
    )
  )

kable(
  x = rmst_metadata_adj_table,
  caption = "TMLE adjusted estimates of Restricted Mean Survival Time.",
  digits = 2
)

kable(
  x = rmst_metadata_unadj_table,
  caption = "TMLE **unadjusted** estimates of Restricted Mean Survival Time.",
  digits = 2
)

```


<p class="forceBreak"></p>


```{r echo = FALSE, warning = FALSE}
survival_probability_adj <-
  adjrct::survprob(
    metadata = surv_metadata_adj,
    horizon = 90
  )

survival_probability_adj_table <-
  with(
    survival_probability_adj$estimates[[1]],

    bind_rows(
      data.frame(
        Arm = "Treatment",
        Estimate = arm1,
        SE = arm1.std.error,
        LCL = arm1.conf.low,
        UCL = arm1.conf.high
      ),

      data.frame(
        Arm = "Control",
        Estimate = arm0,
        SE = arm0.std.error,
        LCL = arm0.conf.low,
        UCL = arm0.conf.high
      ),
      data.frame(
        Arm = "Treatment - Control",
        Estimate = theta,
        SE = std.error,
        LCL = theta.conf.low,
        UCL = theta.conf.high
      )
    )
  )

kable(
  x = survival_probability_adj_table,
  caption = "TMLE adjusted estimates of 90-day Survival.",
  digits = 2
)

kable(
  x = survival_probability_unadj_table,
  caption = "TMLE **unadjusted** estimates of 90-day Survival.",
  digits = 2
)
```




## Summary:

  - Limitations of Logrank & Cox:
    - PH Assumptions; Relative; Dependence on censoring
  - Survival Probability & RMST:
    - Absolute: specific reference to time; Sidestep PH assumption
    - Doubly robust estimators
    - Sample Size Calculations
  - Tutorials: Continuous, Binary, Ordinal, Time-to-Event
    - Example Data: Stroke, Substance Abuse - More forthcoming
    - Fixed Sample Size; Group Sequential; Information Adaptive




## References {.smaller}